\documentclass{article}
\usepackage{graphicx} % immagini
\usepackage{amsmath}  % formule
\usepackage{verbatim} % per includere testo puro
\usepackage{fancyhdr} % eventuale formattazione extra
\usepackage{float}
\usepackage{booktabs}
\usepackage{subcaption}
\usepackage{multirow}
\usepackage[inline]{enumitem}
\usepackage{pgfplots}
\pgfplotsset{compat=1.18}

\date{September 2025}

\begin{document}
    \begin{titlepage}
        \begin{center}
            {\LARGE STS - Sport Tournament Scheduling }
            \vspace*{1em}
            
            Samuele Centanni, Tomaž Cotič, Mattia Lodi, Artem Uskov

            \centerline{\{samuele.centanni, tomaz.cotic, mattia.lodi, artem.uskov\}@studio.unibo.it}
        \end{center}
    \end{titlepage}

\input{Introduction}  
\input{CP}
\input{SAT}     
\input{SMT}    
\input{MIP}
\input{Conclusion}

\section*{Authenticity and Author Contribution Statement}

We declare that this work is our own and not copied from anyone else. The work has been roughly split in the following way: Cotič did the CP part, Centanni worked on SAT, Uskov did the SMT part and Lodi completed the MIP part.

\begin{thebibliography}{9}

\bibitem{dewerra1999}
Dominique de Werra.
\newblock Scheduling Round-Robin Tournaments: An Overview.
\newblock \emph{Discrete Applied Mathematics}, 91(1–3):241--277, 1999.

\bibitem{bailleux2003}
Olivier Bailleux and Yacine Boufkhad.
\newblock Efficient CNF Encoding of Boolean Cardinality Constraints.
\newblock In: \emph{Principles and Practice of Constraint Programming (CP 2003)}, Lecture Notes in Computer Science, 2003.
\end{thebibliography}

\end{document}
